---
title: "Butterfly Model Fit Analysis"
output: 
  html_document: 
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Running models


```{r warning = FALSE, message = FALSE}
library( tidyverse)
library(ggtern)
library(HMSC)
select <- dplyr::select
library(corrplot)
library(broom)
library(cowplot)
```

## Loading Data

Fitting handled by output of `BF_Version2_perc2_log_env.R` i.e. cutting off those with mean less than 2% presence / absence, and logging some of the environmental variables. 

```{r}
Outs<-list.files(path = paste0('../SavedObjects/HMSC_Bf2_perc2_log_env/'), full.names = TRUE)
AllRuns_HMSC2test<-map_df( Outs, read_csv, show_col_types = FALSE)
ButterflyNames <- read_csv('../Data/ButterflyNames.csv', show_col_types = FALSE)

SITES <- read_csv('../Data/Site_selection2_CHALK.csv')

colnames(AllRuns_HMSC2test)
load('../Data/YearSlices2000x2020_v2_2perc_nomoths')   
```

# Correlation 

```{r}
cor(SITES[,-1]) %>% corrplot()
SITES$CHALK %>% table
```

# Occupancy through time

```{r}
map_df(Year_Slices, .f = function( x){
  return(colSums(x[,-1]))
} ) -> OccupancyPerYear


data.frame(TotalOccupancy = 100*rowSums(OccupancyPerYear)/(42*97),
           Year = 2000:2020) %>%
  ggplot( aes( x = Year, y = TotalOccupancy))+
  geom_point()+
  geom_line()+
  theme_minimal()+ ylab( 'Total Occupancy\nPercentage')+
  coord_cartesian(ylim = c(0,100 ))

ggsave( '../ButterflyMarkdowns/ButterflySIFigs/TotalOccupancy.png',
        height = 4, width = 4, dpi = 500)


OccupancyPerYear%>%
  mutate(Year = 2000:2020) %>%
  gather( 'SciName', 'Occupancy', -Year) %>%
  left_join( ButterflyNames, by = "SciName") -> OccupancyDF 
```

### Species Occupancy

```{r}

OccupancyDF %>%
  ggplot(aes( x = Year, y = Occupancy))+
  geom_point()+
  geom_line()+
  facet_wrap(~CommName, ncol = 6)+
  theme_minimal()

ggsave('ButterflySIFigs/SpOccupancy.png',
       height = 10, width = 14, dpi = 500)

```

# Partitioning

## Overall Averages

```{r}
AllRuns_HMSC2test%>%
  group_by(Year) %>%
  summarise(MeanEnv = mean(Part_env), 
            MeanSpa = mean(Part_spa), 
            MeanCoDist = mean(Part_codist),
            MeanR2 = mean(Part_R2_sum)) -> AllRuns_Means_HMSC_cutdown

AllRuns_Means_HMSC_cutdown %>%
  select( Year, Envrionmental =  MeanEnv,
          Spatial =  MeanSpa,
          Codistribution= MeanCoDist) %>%
  gather('Metric', 'Value', -Year  )%>%
  mutate( LineType = ifelse(Metric == 'Codistribution', 'A', 'B' ))%>%
  ggplot( aes( x = Year, y = Value   ))+
  geom_point()+
  facet_wrap(~Metric, scale =  'free_y', nrow = 1)+
  geom_smooth(method = 'lm', se = FALSE, aes( linetype= LineType))+
  ylab(bquote('Variance Explained\n'~(R[adjust]^2)))+ guides( linetype = FALSE)+
  theme_minimal()

ggsave(filename = '../ButterflyMarkdowns/ButterflySIFigs/MainButterflyResults.png',
       width = 10, height = 4, dpi = 500)

```

### Statistics

```{r}
lm(data = AllRuns_Means_HMSC_cutdown, MeanEnv ~ Year) %>% summary
lm(data = AllRuns_Means_HMSC_cutdown, MeanCoDist ~ Year) %>% summary
lm(data = AllRuns_Means_HMSC_cutdown, MeanSpa ~ Year) %>% summary
lm(data = AllRuns_Means_HMSC_cutdown, MeanR2 ~ Year) %>% summary

```
## Split by Species

```{r fig.height=16, fig.width=10}

AllRuns_HMSC2test%>%
  select( Species, Year,
          Environmental = Part_env ,
          Spatial = Part_spa ,
          Codistribution = Part_codist ) %>%
  left_join(OccupancyDF, by = c('Species' = 'SciName', 'Year' )) %>%
  mutate( Occupancy = Occupancy /97) %>%
  gather('Metric', 'Value',    Environmental ,Spatial ,Codistribution, Occupancy )%>%
  ggplot( aes( x = Year, y = Value    , col = Metric   ))+
  geom_point()+
  theme_minimal()+
  theme(legend.position = 'bottom')+
  facet_wrap(~CommName, ncol = 6)+
  scale_colour_manual( values = c('lightskyblue3', 'palegreen4','gold2',  'brown'),
                       limits = c('Codistribution','Environmental' ,'Spatial', 'Occupancy'))+
  geom_smooth(method = 'lm', se = FALSE)

ggsave( '../ButterflyMarkdowns/ButterflySIFigs/SpeciesResponses.png',
        height = 14, width = 12, dpi = 500)

```


# Examining Model Fits

NB full model fits not included in public repository (500Mb)

```{R}
Models<-list.files('../SavedObjects/HMSC_Bf2_perc2_log_env_fits/', full.names = TRUE)

MainModels<- Models[!str_ends(Models, '_2')]
N_Species <- length(unique(AllRuns_HMSC2test$Species))

EnvCoeffs <- array( NA, dim = c( N_Species, 7, 21))  ## [species , var, year]
Omegas <- array( NA, dim = c( N_Species, N_Species, 21))  ## [species , species, year]

for( i in 1:21 ){
  load( MainModels[i])
  fittedParams <-apply( model$results$estimation$paramX,MARGIN =c(1,2), FUN = mean)
  EnvCoeffs[,,i] <- fittedParams[,1:7]
  Array_Omega <-corRandomEff(model, cor = TRUE, burning = FALSE)[,,,1] ##  [sp1, sp2, mcmc slice, latentname]
  Mean_Omega <- apply(Array_Omega, MARGIN = c(1,2), mean)
  Omegas[,,i] <- Mean_Omega
}
```



## Species Envrionmental Dependencies

```{r fig.height=15, fig.width=10}

data.frame( SciName=  colnames(model$data$Y)) %>%
  left_join(ButterflyNames, by = "SciName")-> Names_Ordered

Mean_EnvCoeffs <- apply(EnvCoeffs, c(1,2), mean)
SD_EnvCoeffs <-  apply(EnvCoeffs, c(1,2), sd)


## Names 
 colnames(model$data$X)[1:7]
 
 NiceNames <- c("Intercept",
                "Mean\nTemperature",
                "Woodland",
                "Farming",
                "Urban",
                "Calcareous\nGrassland",
                "Chalk\nBedrock")     
 
colnames(Mean_EnvCoeffs)  <- NiceNames

bind_cols(Mean_EnvCoeffs,Names_Ordered)%>%
  as.data.frame() %>%
  select(-SciName, - Intercept) %>%
  gather('predictor', 'coeff',  - CommName   )%>%
  ggplot(aes( x = predictor, y = CommName, fill = coeff ))+
  geom_tile(alpha = 1)+
  scale_fill_gradient2(mid = 0, name = 'Average\nFitted\nCoefficient')+
    scale_y_discrete(limits=rev)+
  ylab('') +xlab('') -> Env_Coeff_Plot

ggsave(Env_Coeff_Plot, 
       filename = '../ButterflyMarkdowns/ButterflySIFigs/EnvCoeffs.png',
       width = 8, height = 10, dpi = 500)
        

###
colnames(SD_EnvCoeffs)  <- NiceNames

bind_cols(SD_EnvCoeffs,Names_Ordered)%>%
  as.data.frame() %>%
  select(-SciName, - Intercept) %>%
  gather('predictor', 'SD_coeff',  - CommName   ) %>%
  ggplot(aes( x = predictor, y = CommName, fill = SD_coeff ))+
  geom_tile(alpha = 1)+
    scale_y_discrete(limits=rev)+
  ylab('') +xlab('') +
  scale_fill_viridis_c() -> Env_SD_Plot


plot_grid(Env_Coeff_Plot+theme(legend.position = 'bottom') ,
          Env_SD_Plot+theme(legend.position = 'bottom'),
          scale = 0.9,
          labels = c('Average fitted coefficent', 
                     'SD in fitted coefficent'))

```


## Species Associations. 

```{r}
Mean_Omegas <- apply(Omegas, c(1,2), mean)
SD_Omegas <-  apply(Omegas, c(1,2), sd)

rownames(Mean_Omegas) <- Names_Ordered$CommName
rownames(SD_Omegas) <- Names_Ordered$CommName
colnames(Mean_Omegas) <- Names_Ordered$CommName
colnames(SD_Omegas) <- Names_Ordered$CommName

```

```{R fig.height=12, fig.width=12}

Mean_Omegas %>%
    corrplot(corr =., title = 'Mean Omega', order = 'hclust', tl.col = 1)

png(filename = '../ButterflyMarkdowns/ButterflySIFigs/SpAssCoeffs.png',
    width = 12, height = 12, res = 500, units = 'in' )

Mean_Omegas %>%
    corrplot(corr =., order = 'hclust', tl.col = 1)
dev.off()


  SD_Omegas %>%
    corrplot(corr =., title = 'SD Omega', order = 'hclust')
```


# Convergence

### Year 2000
```{r eval = FALSE}

library(coda)

SpreadArray <- function( i, Array_Omega){ 
  ##for coda, need each element in a separate column, but only need lower half of diagonal
  XX<- Array_Omega[,,i,1]
  return(as.data.frame(t(as.matrix(XX[lower.tri(XX)], nrow =1))))
}

CalcGelmanDiags <- function(Year){
  
  
  load(paste0('../SavedObjects/HMSC_Bf2_perc2_log_env_fits/model_',Year))
  load(paste0('../SavedObjects/HMSC_Bf2_perc2_log_env_fits/model_',Year,'_2'))
  
  ### Main predictors (Spatial and Envrionmental)
  coda1_param <- as.mcmc(model,  parameters = "paramX")
  coda2_param <- as.mcmc(model2, parameters = "paramX")
  
  coda_list = mcmc.list(coda1_param,coda2_param)
  
  Gelmans_Main<- gelman.diag(coda_list, multivariate = FALSE)
  
  ## Convergence of latents
  ## a bit more complex, as need to first get to 'omegas',  the residual effect of each species on each other.  
  
  Array_Omega1 <-corRandomEff(model, cor = TRUE, burning = FALSE) ##  [sp1, sp2, mcmc slice, latentname]
  Array_Omega2 <-corRandomEff(model2, cor = TRUE, burning = FALSE) ##  [sp1, sp2, mcmc slice, latentname]
  
  MCMC_length = dim(Array_Omega1)[3]
  Omega_MCMC1<- map_dfr(1:MCMC_length, SpreadArray, Array_Omega1)
  Omega_MCMC2<- map_dfr(1:MCMC_length, SpreadArray, Array_Omega2)
  
  Omega_MCMC_list <- mcmc.list(as.mcmc(Omega_MCMC1), as.mcmc(Omega_MCMC2))

  Gelmans_omega<- gelman.diag(Omega_MCMC_list, multivariate = FALSE, transform = TRUE)
  
  Gelmans_omega_df <- as.data.frame(Gelmans_omega$psrf) 
  Gelmans_Main_df <- as.data.frame(Gelmans_Main$psrf)
  Gelmans_omega_df$Year <-Year
  Gelmans_Main_df$Year <-Year
  cat(Year)
  gc()
  return(list( Main =Gelmans_Main_df, Omega =Gelmans_omega_df  ) )
}


GelmanDiagResults<- purrr::map(2000:2020, CalcGelmanDiags)

### Takes a little while, so saving
save(GelmanDiagResults, file = '../SavedObjects/GelmanDiagResults_logenvfits' )
```


```{r}

load('../SavedObjects/GelmanDiagResults_logenvfits' )

MainCoefficents <- map_df(GelmanDiagResults,  'Main')
Omegas <- map_df(GelmanDiagResults,  'Omega')


MainCoefficents %>%
  ggplot(aes( `Point est.`))+
  geom_histogram(bins = 50)+
 # facet_wrap(~Year)+
  geom_vline( xintercept = 1.1, linetype= 3)+
    geom_vline( xintercept =1.2, linetype= 2)+
  theme_minimal()+
  xlab('Point Estimate of\nGelman-Rubin Diagnostic') -> MainEffects_plot


Omegas %>%
  ggplot(aes( `Point est.`))+
  geom_histogram(bins = 50)+
  geom_vline( xintercept = 1.1, linetype= 3)+
    geom_vline( xintercept =1.2, linetype= 2)+
  theme_minimal()+
  xlab('Point Estimate of\nGelman-Rubin Diagnostic') -> Omega_plot


Omegas%>%
  arrange(desc( `Point est.`)) %>%
  head



plot_grid(MainEffects_plot, Omega_plot, 
          labels = c('a) Main Effects',
                     'b) Omega Entries' ), scale = 0.9)

ggsave('ButterflySIFigs/ButterflyConvergence_GR.png', 
       width = 10, height = 5, dpi = 500)

```


# Map of Sites

```{r}

library(maps)  # nb this overwrites purrr ,map()
UK =  map_data("world") %>% filter(region == 'UK')

SiteDetails <- read_csv('../Data/SiteDetails.csv') %>%
  select( OSGR, Lat, Long)

SITES %>%
  left_join( SiteDetails, by = 'OSGR') %>%
  ggplot( aes( x = Long, y = Lat))+
  geom_map(data = UK, map = UK, aes(long, lat, map_id = region) , fill = 'grey', col = 'grey25')+
  geom_point(pch = 4, size = 0.8)+
  scale_color_viridis_d()+
  geom_hline(yintercept = 53.5)+
  geom_vline(xintercept = -3.5)+
  theme_minimal()+
  coord_map(xlim = c( -4, 2), ylim = c( 50, 55))+
  ylab('')+
  xlab('')

ggsave( 'ButterflySIFigs/ButterflySiteMaps.png', height = 5, width = 4, dpi = 500)

```

