---
title: "Butterfly Data Preparation"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This document summarises the data processing to the butterfly observation and transect environmental data. NB many of the larger input files are not included as they too large, but I point to their sources. This document may not run nicely, but should document all the steps carried out. 

 
```{r message = FALSE, warning=FALSE}
library(tidyverse)
select <- dplyr::select
```

# Butterfly observation data

Original .csv's of butterfly observations were downloaded from https://records.nbnatlas.org/ covering 1970 - 2020. These were then combined into`UKBMS_JoinedAllYears.csv` (too large for Github). 

```{r eval = FALSE}
UKBMS_files <- list.files('D:/UKBMS/', full.names = TRUE  )  #  location of downloads from https://records.nbnatlas.org/ # (NB limit of 500'000  per batch)

JoinedAllYears <-map_df(UKBMS_files, function(file){
  
  read_csv(file, col_select = c(`Scientific name`,
                                `Start date year`,
                                OSGR, `State/Province` , 
                                `Latitude (WGS84)`,
                                `Longitude (WGS84)`)) %>%
    select( SciName = `Scientific name`,
            Year =`Start date year`,
            OSGR, `State/Province` , 
            Lat = `Latitude (WGS84)`,
            Long = `Longitude (WGS84)`) %>%
    filter( `State/Province` == 'England')  %>%
    select(- `State/Province`  )-> xx
  return(xx)
})

# write_csv(JoinedAllYears, 'UKBMS_JoinedAllYears.csv')
```

```{r}
UKBMS <- read_csv('../../Butterflydata/Data forButterflies/UKBMS_JoinedAllYears.csv') ## now has 1990-2020 in there
## includes still multiple observations from the same site/year (i.e different dates)

Sites <- read_csv('../../Butterflydata/Data forButterflies/SiteDetails.csv')   #
ButterflyNames <- read_csv('../ButterflyNames.csv')

UKBMS_JoinedAllYears%>%
  select(  OSGR,Lat , Long) %>%
  distinct( OSGR,Lat,Long) -> SiteDetails

```

# Envrionmental Predictors

## Temperature

Source and Citation: https://rmets.onlinelibrary.wiley.com/doi/full/10.1002/gdj3.78

```{r eval = FALSE}
nc.brick <- brick('tas_hadukgrid_uk_5km_ann-30y_196101-199012.nc') ## too large for github
nc.df <- as.data.frame(nc.brick[[1]], xy=TRUE)
# write_csv(nc.df, 'tas_hadukgrid_uk_5km_ann-30y_196101-199012.csv')
```

```{r}
MeanAnnualTemps60to90 <- read_csv('tas_hadukgrid_uk_5km_ann-30y_196101-199012.csv')

RasterProjection_Temps<- '+proj=tmerc +pm=0 +a=6377563.396 +b=6356256.909 +lon_0=-2 +lat_0=49 +x_0=4e+05 +y_0=-1e+05 +k_0=0.9996012717'

### Getting data for all sites

SiteDetails %>%
  select(Long,Lat )%>%
  as.matrix() %>%
  project(proj = RasterProjection_Temps)%>%
  raster::cellFromXY(object = nc.brick,xy = .) -> Cells_MeanTemps

SiteDetails$CellID_MeanTemps <- Cells_MeanTemps
SiteDetails$MeanTemp <-raster::extract(nc.brick[[1]],SiteDetails$CellID_MeanTemps)
```


## Landcover data

Source: 
https://catalogue.ceh.ac.uk/documents/d5ee5360-12c5-448c-9d2b-f5c941fe3948

Fuller, R.M.; Smith, G.M.; Sanderson J.M. ; Hill, R.A.; Thomson, A.G.; Cox, R.; Brown, N.J.; Clarke, R.T.; Rothery, P.; Gerard, F.F. (2002). Land Cover Map 2000 (1km percentage target class, GB). NERC Environmental Information Data Centre. https://doi.org/10.5285/d5ee5360-12c5-448c-9d2b-f5c941fe3948 


```{r}
LC2000 <- brick( 'CEHLandcover2000/cover_2000/data/LCM2020/LCM2000_GB_1K_PC_TAR.tif')
str(LC2000)

RasterProjection<- '+proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717 +x_0=400000 +y_0=-100000 +ellps=airy +towgs84=446.448,-125.157,542.06,0.15,0.247,0.842,-20.489 +units=m +no_defs'

```

####  Land cover classes

from the metadata:
```{r}
LCM_habitats <- c('Sea / Estuary',
'Water (inland)',
'Littoral rock',
'Littoral sediment',
'Saltmarsh',
'Supra-littoral rock',
'Supra-littoral sediment',
'Bog (deep peat) ',
'Dense dwarf shrub heath',
'Open dwarf shrub heath',
'Montane habitats',
'Broad-leaved / mixed woodland ',
'Coniferous woodland',
'Improved grassland',
'Neutral grassland',
'Setaside grassland',
'Bracken',
'Calcareous grassland',
'Acid grassland',
'Fen, marsh, swamp',
'Arable cereals',
'Arable horticulture',
'Arable non-rotational',
'Suburban / rural developed',
'Continuous urban',
'Inland bare ground')
```


```{r}
SiteDetails %>%
  select(Long,Lat )%>%
  as.matrix%>%
  project(proj = RasterProjection)%>%
  raster::cellFromXY(object = LC2000,xy = .) -> Cells_LCM

SiteDetails$CellIDs_LCM <- Cells_LCM

# container to store the various land covers
AllHabitatValues<- matrix(NA,nrow= length(Cells_LCM), ncol= 26) 

colnames(AllHabitatValues) <- LCM_habitats

for(i in 1:length(LCM_habitats)){
  AllHabitatValues[,i]<-raster::extract(LC2000[[i]], Cells_LCM)
}

SiteDetails <- bind_cols(SiteDetails,as.data.frame( AllHabitatValues))
# write_csv(SiteDetails, 'SiteDetails.csv' )

```

# Identifying Sites

## Number of Years Visited + Walks per Year

```{r}
YearToStart = 2000

UKBMS %>%
  filter( Year>= YearToStart) %>%
  distinct(OSGR, Year) %>%
  count(OSGR, name = 'NumYears') -> Sites_n_Years

#### Number of years visited at least 5 times
UKBMS %>%
  distinct(OSGR, Year,Month, Day) %>%
  arrange(OSGR, Year ) %>%
  count(OSGR, Year, name = 'NumDaysInYear') -> DaysWalked_SiteYear

DaysWalked_SiteYear %>%
    filter( Year>= YearToStart) %>%
  group_by(OSGR) %>%
  summarise(NumYearsAtLeast5 = sum(NumDaysInYear>=5),
            FirstYear = min(Year), 
            LastYear = max(Year)) %>%
  left_join( Sites_n_Years , by = "OSGR") -> Sites_n_Years_5times

```

### Identifying which years are missing:

```{r}

DaysWalked_SiteYear %>%
  filter( Year>= YearToStart) %>%
  spread( Year,NumDaysInYear )  %>%
  gather( 'Year','NumDaysInYear', -OSGR  )%>%
  filter( is.na(NumDaysInYear )) %>%
  arrange( OSGR, Year) %>%
  mutate( Year = as.numeric( Year), 
          DoubleGap =  (Year - lag(Year))==1 ) %>%
  group_by(OSGR) %>%
  summarise(Num_DoubleGaps = sum(DoubleGap, na.rm=TRUE))  %>%
  left_join(Sites_n_Years_5times , by = "OSGR") -> sample_completeness_data

sample_completeness_data

```

## Filtering down sites based on location and data completeness

## Envrionmental Data selection 

```{r}
Sites %>%
  transmute(OSGR = OSGR,Lat=Lat,Long=Long,
          MeanTemp = MeanTemp, 
         Woodland = `Broad-leaved / mixed woodland` +  `Coniferous woodland`,
          Farming = `Arable horticulture` +  `Arable cereals` +  `Improved grassland` + `Arable non-rotational` ,
          Urban =  `Continuous urban` + `Suburban / rural developed`,
          Calc_Grass = `Calcareous grassland`) -> Sites_simpl
```


```{r}
Sites_simpl %>%
  filter( Long>-3.5, Lat <53.5) %>% # English squarish box, to remove peripheral sites in north and Cornwall/Devon
  left_join(sample_completeness_data , by = 'OSGR') %>%
  filter(!is.na(NumYears)) %>% # this gets rid of sites only sampled pre 1990
  filter( !is.na(MeanTemp )) %>% ## problems of not getting good temperature measures on coast 
  filter(NumYearsAtLeast5 >17) %>% # have at most 3 gaps
  filter(LastYear >= 2019 ) %>%  # don't end on a double gap
  filter(FirstYear <=2001 ) %>%  # don't start on a double gap
  filter( str_length(OSGR)==6 ) %>%  # remove vague grid references ## ST43 linked to multiple Lat_longs 
  arrange(desc(Num_DoubleGaps)) -> SiteSelection


SiteDetails <- SiteSelection
```

# Chalk data 

https://www.bgs.ac.uk/?wpdmdl=12059

Terms used:

*High* Parent type has a high CaCO 3 content due to its primary mineralogy e.g. Chalk.
*Variable(high)* Parent type has a variable but possibly high CaCO 3 content due to lithological variation e.g. Interbedded limestone and calcareous mudstone beds.
*Mod* Parent type has a moderate CaCO 3 content due to its primary mineralogy e.g. dolomite.
*Variable* Parent type has a variable (high to low) CaCO3 content due to heterolithic distribution of carbonate e.g. Till with chalk-gravel.
*Low* Parent type has a low CaCO 3 content due to its primary mineralogy e.g. weakly calcareous-cemented mudstone.
*Variable(low)* Parent type has a variable but typically low CaCO 3 content e.g. mudstone with intermittent, weakly-calcareous, sandstone beds.
*None* Parent type has a no primary CaCO 3 content.
*Unknown* Parent materials whose precise lithology is unclear (generally highly variable at short scales) or from a location that no mineralogical/geochemical data is available to confirm primary mineralogical content.
*Not applicable* No data applicable (typically inland water bodies).

```{r}
ParentMaterial <- read_sf('D:/SPMM_1km/SoilParentMateriall_V1_portal1km.shp')  ## too large for github

CRS<-st_crs(ParentMaterial)

Carbonate <- select( ParentMaterial, CARB_CNTNT) 

Carbonate %>%
  mutate( HighChalk = CARB_CNTNT %in% c( 'HIGH' ,'VARIABLE(HIGH)' )) ->Carbonate 

Focal_xy<-  as.matrix(select(SiteDetails, Long, Lat))

# WGS84 to BNG
Sites_sf <- st_as_sf(as.data.frame(Focal_xy), coords = c("Long","Lat"),crs=4326)
Sites_sf_27700 <- DT1_sf %>% st_transform(crs = CRS)  


st_intersection(DT1_sf_27700,Carbonate,  sparse = FALSE) -> Site_Carb_CNTNT

Draws <- sample( 1:nrow(Carbonate), size = 20000, replace = FALSE)

SiteDetails2$CHALK <- Site_Carb_CNTNT$HighChalk


SiteDetails2%>%
  select(  OSGR ,MeanTemp ,Woodland ,Farming ,Urban, Calc_Grass, CHALK) #%>%
#write_csv('~/GitHub/MC_coherence/Site_selection2_CHALK.csv')

```


# MEM for selected sites

```{r}
SiteDetails2 <- read_csv('~/GitHub/MC_coherence/Site_selection2_CHALK.csv')
Focal_xy<-  as.matrix(select(SiteDetails2, Long, Lat))
```

```{r eval = FALSE}
MEM <- dbmem(Focal_xy, MEM.autocor = "positive", silent = FALSE)

# write_csv(bind_cols(data.frame(OSGR=SiteDetails2$OSGR),
                    MEM[,1:10]),
          '~/GitHub/MC_coherence/MEM_justfocalsites_first10.csv')
```



# Selecting Species

## Species frequencies in selected sites

```{r}
UKBMS  %>%
  filter( Year >YearToStart-1) %>%
  filter(OSGR %in% SiteSelection$OSGR) %>%
  distinct( SciName, Year, OSGR   )%>%
  mutate( Present = 1) -> UKBMS_selected_sites_yearly 

UKBMS_selected_sites_yearly %>%
  distinct(Year, OSGR) %>% nrow  -> TotalSiteYears ### 1885

UKBMS_selected_sites_yearly %>%
  count( SciName, Year) %>%
  group_by(SciName) %>%
  summarise(Min_N = min(n), Max_N = max(n), 
            Mean_N = mean(n), Overall_Perc_Occ = 100* sum(n)/TotalSiteYears) %>%
  arrange(Overall_Perc_Occ) %>%
  left_join(ButterflyNames, by = "SciName") -> Species_Frequencies

Species_Frequencies %>% as.data.frame()

```

## Moths \ Vagrants

Burnet Companion (*Euclidia glyphica*) and Cinnabar (*Tyria jacobaeae*) are both moths and therefore not necessarily counted on all transects. 
 
Clouded yellow (*Colias croceus*) and Painted lady (*Vanessa cardui*) both migrants from Africa - largely unconnected dynamics as even though they can get a generation or two in, they don't survivie the winter.  

```{r}
Species_Frequencies %>%
  filter(Overall_Perc_Occ >= 2 & Overall_Perc_Occ <= 98  ) %>%  # Average at least 2% presences or absences
  filter( !(SciName %in%c('Tyria jacobaeae', 'Euclidia glyphica',
                          'Colias croceus',
                          'Vanessa cardui')  )) %>%  # exclude moths and seasonal migrants
pull(SciName)-> SpeciesToUse

```


# Inferring gaps in data. 

```{r}
UKBMS_selected_sites_yearly %>%
  filter(SciName %in% SpeciesToUse) -> 
  UKBMS_selected_sites_yearly_focal

### Adding zeros where not present, and NA's when no walks done

UKBMS_selected_sites_yearly_focal %>%
  arrange( SciName)%>%
  pivot_wider( id_cols = c( OSGR,Year   ),    ## This first flex adds 0s where there was a survey, but a apseices wasn't seen
               names_from = SciName,
               values_from = Present,
               values_fill = 0) %>%
  pivot_longer( names_to ='SciName' ,
                values_to = 'Present',
                cols = !c( OSGR,Year) ) %>%
  arrange( Year)%>%
  pivot_wider( id_cols = c( OSGR, SciName  ), # This second flex adds NAs to species columns where there was no survey in that year
               names_from = Year,
               values_from = Present,
               values_fill = NA)  -> UKBMS_selected_sites_wide_Zero_or_NA

```

### Fraction of gaps

```{r}
UKBMS_selected_sites_wide_Zero_or_NA %>%
  group_by(OSGR) %>%
 pivot_longer(names_to ='Year' ,
                values_to = 'Present',
                cols = !c( OSGR,SciName)) %>%
  group_by(OSGR, Year) %>%
  summarise(N_sp = sum(Present)) %>%
  ungroup() -> Gaps

Gaps%>%
  summarise(OverallGapsPerTimeSeries = sum( is.na(N_sp))/97)

```


### function to fill in gaps
To keep the number of sites consistent, will infer the gaps based on close time points. 

Infer gaps based on neighbours. Need some rules:
- if present before and after, then fill 
- if not present before and after, then don't fill
- if present on one side, fill with probability = occurence frequency. 

this cycles through each site and fills in gaps as need:

Kind of a horrible from a coding perspective, but fast enough for the job...

```{r}
SiteVec<- SiteSelection$OSGR

 GAP_FILLER <- function(i, SiteVec){
  
  UKBMS_selected_sites_wide_Zero_or_NA %>%
    filter( OSGR == SiteVec[i]) %>%
    select( -OSGR, -SciName) %>%
    as.matrix-> FocalSite
  
  FocalSite_ToFill = FocalSite
  
  if(any(is.na( FocalSite))){  ## see if there are gaps that need filling
    
    ## find gaps 
    gaps<-   which(is.na(FocalSite[1,]))
    filled <-  which(!is.na(FocalSite[1,]))
    
    for( g in gaps)  {
      ## Can get cases where there is no previous or later information if gaps are at end. 
      ## In that case, just use what information we have
      closest_before<-NA ;  closest_after <-NA
      
      try({if( (g-1) %in% filled){closest_before= g-1}else{
        if( (g-2) %in% filled){closest_before= g-2}else{
          if( (g-3) %in% filled){closest_before= g-3}  }  }  })
      try({ if( (g+1) %in% filled){closest_after= g+1}else{
        if( (g+2) %in% filled){closest_after= g+2}else{
          if( (g+3) %in% filled){closest_after= g+3}  }  }})
      
      for( sp in 1: nrow(FocalSite )){
        
        ToFill <- mean(c(FocalSite[sp,closest_before], FocalSite[sp,closest_after]),
                       na.rm = TRUE)  # c(1,  NA), or c( 0,NA) will give 0 or 1, making ends ok
        
        # if disagreement, (0,1) or (1,0) will return 0.5. fill with a draw based on overall occurence
        if(ToFill ==0.5 ){
          ToFill = rbernoulli(1, p = mean(FocalSite[sp,], na.rm=TRUE))
        }
        FocalSite_ToFill[sp,g] <- ToFill
      }
    }
  }
  bind_cols(  UKBMS_selected_sites_wide_Zero_or_NA %>%
                filter( OSGR == SiteVec[i]) %>%
                select( OSGR, SciName), 
              FocalSite_ToFill)%>%
   # mutate(FilledGaps = paste(gaps, collapse = '_')) %>% 
    return()
}
set.seed(1)
UKBMS_selected_sites_FilledGaps<- map_df(1:length(SiteVec),GAP_FILLER,SiteVec)
```

## Spreading into a list object

```{r}
Years = colnames(UKBMS_selected_sites_FilledGaps)[-c(1,2)]

 SPREAD_TO_LIST <- function( i, UKBMS_selected_sites_FilledGaps){
  
  UKBMS_selected_sites_FilledGaps[,c(1,2,i+2)] %>%
    spread(SciName, value = 3) -> Site_x_Species
  return(Site_x_Species )
}
 
Year_Slices <- map( 1:length(Years),SPREAD_TO_LIST,UKBMS_selected_sites_FilledGaps)

names(Year_Slices) <- Years

save(Year_Slices, file = '../Data/YearSlices2000x2020_v2_2perc_nomoths')  
write_csv( UKBMS_selected_sites_FilledGaps, '../Data/UKBMS_selected_sites_FilledGaps_v2_2perc_nomoths.csv')
```
