---
title: "Birds Data Preparation"
output: 
  html_document: 
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r message = FALSE, warning= FALSE}
library(tidyverse)
select<- dplyr::select
map<-purrr::map


```

```{r message = FALSE}
## Data from https://www.bto.org/sites/default/files/atlas_open_data_files.zip

Distributions <- read_csv('../atlas_open_data_files/distributions.csv')
GridCells <- read_csv('../atlas_open_data_files/grid_square_coordinates_lookup.csv')

Species <- read_csv('../atlas_open_data_files/species_lookup.csv')

Distributions$period %>% table
Distributions$status %>% table

```

# Bird Data
## Grid cells

```{r eval = FALSE}
## Grid Cell location:
#5 values given are the four corners of the bounding boxes (1st one is repeated)
#To simplify, take an average of first 4
GridCells %>%
  group_by(grid) %>%
  filter( order != 5) %>%
  summarise(long = mean(long),
            lat = mean(lat)) -> GridCenters
```

## Filtering bird records to final set

```{r eval = FALSE}
## 
Distributions %>%
  filter( season == 'B',
          period %in% c('1968-72', '2008-11'),
          island  == 'B', 
          resolution == 10) %>%
  left_join(GridCenters, by ="grid") %>%
    select( -season, - island, - resolution) %>%
  left_join(Species, by = "speccode") %>%
  filter( lat>= 51, lat <= 54.3,
          long >= -2.3, long <=-0.9 )  %>%
  add_count(english_name, name = 'total_count') %>%
  filter( total_count >=16,
          total_count < 680) ->CentralBreeders_focal
```

## Saving Species data as presence/absence matrices
```{r eval = FALSE}
## Higher certainty 
CentralBreeders_focal %>%
  filter( status %in% c('Confirmed', 'Probable')) %>%
  select( period, grid,  english_name) %>% 
  mutate(Present = 1) %>%
  pivot_wider(id_cols = c(period, english_name),
              names_from = grid,
              values_from = Present ,
              values_fill = 0) -> HigherCert

## All data certainty 
CentralBreeders_focal %>%
  select( period, grid,  english_name) %>% 
  mutate(Present = 1) %>%
  pivot_wider(id_cols = c(period, english_name),
              names_from = grid,
              values_from = Present ,
              values_fill = 0) -> AllCert

## Saving separately
HigherCert %>% filter( period == '1968-72') %>% select(-period) %>%
  write_csv('../BirdData_clean/focalBirds_High_1970.csv')
HigherCert %>% filter( period == '2008-11') %>% select(-period) %>%
  write_csv('../BirdData_clean/focalBirds_High_2010.csv')
AllCert    %>% filter( period == '1968-72') %>% select(-period) %>%
  write_csv('../BirdData_clean/focalBirds_AllC_1970.csv')
AllCert    %>% filter( period == '2008-11') %>% select(-period) %>%
  write_csv('../BirdData_clean/focalBirds_AllC_2010.csv')
```

# Envrionmental data

## Land cover

Land cover map available for 1990 - intermediate between samples.
https://catalogue.ceh.ac.uk/documents/bb381b5b-d44e-4dbd-a9d1-efffd4c3e4a8

```{r eval = FALSE}
library(raster)
library(rgdal)

LC1990 <- brick( 'D:/Data For Birds/Land Cover Map 1990 1km perc/data/gb1990lcm1km_pc.tif')
# str(LC1990)

RasterProjection<- '+proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717 +x_0=400000 +y_0=-100000 +ellps=airy +towgs84=446.448,-125.157,542.06,0.15,0.247,0.842,-20.489 +units=m +no_defs'

LCM_habitats_1990 <- c( 'Broadleaved woodland', 'Coniferous Woodland','Arable and Horticulture',
                        'Improved Grassland','Neutral Grassland','Calcareous Grassland','Acid grassland','Fen, Marsh and Swamp','Heather',
                        'Heather grassland','Bog','Inland Rock','Saltwater','Freshwater','Supra-littoral Rock',
                        'Supra-littoral Sediment','Littoral Rock','Littoral sediment',
                        'Saltmarsh','Urban','Suburban')
```

```{r eval =FALSE}
CentralBreeders_focal %>%
  distinct( long, lat, grid) -> Focal_grid_coords

GridCells %>%
  filter( grid %in% Focal_grid_coords$grid,
          order !=5 ) -> CellEdges  # this gives the corners of each grid cell. 

Find_Average_LandCover <- function( FG){
  
  Edges<- filter( CellEdges,grid == FG)
  
  edge_lats = range(Edges$lat)
  edge_lons = range(Edges$long)
  
  mid_lats = seq( from = edge_lats[1],to = edge_lats[2], 
                  length = 21)[seq(2, 20, by =2)]  
  
  mid_longs = seq( from = edge_lons[1],to = edge_lons[2], 
                   length = 21)[seq(2, 20, by =2)]  
  
  # 100 mid points of each of the 1km grid cells with the 10x10 # should align nicely
  expand.grid(mid_longs, mid_lats )%>% 
    as.matrix() %>%
    project(proj = RasterProjection)%>%
    raster::cellFromXY(object = LC1990,xy = .) -> Cells_LCM  
  Onekm_by_LCvar <- raster::extract(LC1990, Cells_LCM)
  Average_LCvar <- as.data.frame(t(colMeans(Onekm_by_LCvar)))
  Average_LCvar$grid = FG
  cat(FG)
  return(Average_LCvar) 
}

```

```{r  eval = FALSE}
FocalGridLandCover <- map_df( Focal_grid_coords$grid, Find_Average_LandCover) # 1 second or so each, a couple of minutes in total. 
colnames(FocalGridLandCover) <- c(LCM_habitats_1990, 'grid')
write_csv(FocalGridLandCover, '../SavedObjects/BirdsFocalGridLandCover.csv' )
```

### Simplifying land cover data

- Excluding all coastal ones
- Grouping some into aggregate classes, as described in the dataset doc

```{r}
FocalGridLandCover <- read_csv('../BirdData_clean/BirdsFocalGridLandCover.csv' )

FocalGridLandCover %>%
  select( -'Supra-littoral Rock', - 'Supra-littoral Sediment',
          -'Littoral Rock',- 'Littoral sediment', 
          - 'Saltwater', - 'Saltmarsh') %>%
  mutate( Mountain_Heath_Bog = Heather+
            `Heather grassland`+
            Bog +`Inland Rock`) %>%
  select ( -Heather,-`Heather grassland`,
           -Bog ,-`Inland Rock`) %>%
  select(grid, everything())-> InlandHabitats

colMeans(InlandHabitats[,-1])
```

# Spatial Associations (MEM pre-calculations)

```{r}
### creating matrix of spatial coordinates:
## return only those linked to positive autocorrelation. (i.e. removing the negative autocorrelation bits)
Focal_xy<-  as.matrix(Focal_grid_coords[,2:3])

```

```{r eval = FALSE}
library(adespatial)
MEM <- adespatial::dbmem(Focal_xy, MEM.autocor = "positive", silent = FALSE)
MEMs10 <- as.data.frame(as.matrix(MEM[,1:10]))
colnames(MEMs10) <- paste0('MEM', 1:10)

Site_MEMs10 <- bind_cols(data.frame(grid=Focal_grid_coords$grid),
                   MEMs10)
        
write_csv(Site_MEMs10, '../BirdData_clean/Bird_FocalSites_MEM10.csv')
```





# Saving Overall Procesesed Data

## Sites
```{r}
Site_MEMs10<- read_csv('../BirdData_clean/Bird_FocalSites_MEM10.csv')

InlandHabitats %>%
  left_join(Site_MEMs10, by = "grid") %>%
  arrange(grid) %>%
  write_csv( '../BirdData_clean/Birds_AllPreds_noELEV.csv')
```

# Data and Package loading

```{r warning = FALSE, message = FALSE}
library(vegan)

source( '../Scripts/variPart_returnboth.R')  ## variPart_returnboth

Predictors <- read_csv('../BirdData_clean/Birds_AllPreds_noELEV.csv' )
PA_1970_all <-  read_csv('../BirdData_clean/focalBirds_AllC_1970.csv' )
PA_1970_HQ  <-  read_csv('../BirdData_clean/focalBirds_High_1970.csv' )
PA_2010_all <-  read_csv('../BirdData_clean/focalBirds_AllC_2010.csv' )
PA_2010_HQ  <-  read_csv('../BirdData_clean/focalBirds_High_2010.csv' )

```

# Predictors to keep

```{r}
corrplot::corrplot(cor(Predictors[,-1]))
## - Very strong corrlations between urban and suburban

corrplot::corrplot(cor(Predictors[,c(4,8, 13, 15, 11, 12)]))

## Combine the urban predictors
Predictors %>%
  arrange(grid) %>%
  mutate( UrbanSuburban = Urban+Suburban) %>%
  select(-grid, - Urban, -Suburban) %>%
select(`Broadleaved woodland`:Mountain_Heath_Bog,UrbanSuburban, starts_with('MEM')) %>%  # reorder
  as.matrix() -> X_formatted  

rownames(X_formatted) <- arrange(Predictors,grid)$grid  

write.csv(X_formatted, '../BirdData_clean/Bird_X_formatted.csv')
```

# Species to keep 


## Not too common or rare 

Set threshold at least 5 sites must be occupied or unoccupied, *in both years*. NB excludes some big changers like Red Kites...

```{r}
### function that keeps only those that fit criteria

InformativeOccu <- function(Occ_mat , minSites = 5){
  n_sites = (ncol(Occ_mat)-1) 
  Occupancy <-  rowSums(Occ_mat[,-1])
  Occ_mat_informative =  Occ_mat[ (Occupancy>=minSites)  & (n_sites-Occupancy >= minSites), ]
  return(Occ_mat_informative)
}

inform_PA_1970_all<- InformativeOccu(PA_1970_all)
inform_PA_1970_HQ <- InformativeOccu(PA_1970_HQ )
inform_PA_2010_all<- InformativeOccu(PA_2010_all)
inform_PA_2010_HQ <- InformativeOccu(PA_2010_HQ )

```

## Appear in both

Making sure that there is a consistent list of species in both trials

```{R}

## High quality 
Shared_HQ = intersect(inform_PA_1970_HQ$english_name,
                      inform_PA_2010_HQ$english_name )

## All 
Shared_all  = intersect(inform_PA_1970_all$english_name,
                        inform_PA_2010_all$english_name )
```

## Excluding non-native oddities

```{r}
bou_LIST <- read_csv('../BirdData_clean/BOU_British_List.csv')
count(bou_LIST, Category)

bou_LIST %>%
  mutate(  CatAorC2 = str_detect(Category, 'A')|str_detect(Category, 'C2'), 
           Introduced = str_detect(Category, 'C')) %>%
  filter(!CatAorC2)


bind_rows(PA_1970_all,PA_2010_all ) %>%
  distinct(english_name) %>%
  left_join( bou_LIST, by = c('english_name' = 'British (English) vernacular name')) -> withBOU_Categories 

## 11 weird unwanted species either not in overlap or not in category A or C2 (Naturalized established species – species with established populations resulting from introduction by Man, but which also occur in an apparently natural state,)

withBOU_Categories%>%
  mutate(  CatAorC2 = str_detect(Category, 'A')|str_detect(Category, 'C2'), 
           Introduced = str_detect(Category, 'C'))%>%
  filter(!CatAorC2 | is.na(Category)) -> SpeciesToExclude_nonnative

SpeciesToExclude_nonnative

```

```{r}
### removing non_natives
## 75 species
finalSP_HQ <- Shared_HQ[!(Shared_HQ %in% SpeciesToExclude_nonnative$english_name)]  ##  "Ruddy Duck""Red-legged Partridge" "Little Owl"  

## 80 species
finalSP_all <- Shared_all[!(Shared_all %in% SpeciesToExclude_nonnative$english_name)]  ##  "Ruddy Duck""Red-legged Partridge" "Little Owl"  


save(finalSP_HQ, file = '../BirdData_clean/finalSP_HQ' )
save(finalSP_all, file = '../BirdData_clean/finalSP_all' )

```



# Plotting a Map of Sites

```{r}
library(maps)
UK =  map_data("world") %>% filter(region == 'UK')

Focal_xy %>%
  as.data.frame() %>%
  ggplot( aes( x = long, y = lat))+
  geom_map(data = UK, map = UK, aes(long, lat, map_id = region) , fill = 'grey')+
  geom_point(shape = 4)+
  scale_color_viridis_d()+
  geom_hline(yintercept =c(51,54.3))+
  geom_vline(xintercept = c(-2.3,-0.9 ))+
  theme_minimal()+
  coord_map(xlim = c( -4, 2), ylim = c( 50, 55))+
  ylab('')+xlab('')
ggsave( 'BirdImages/HectadLocations.png', height = 5, width = 4, dpi = 500)
```
