---
title: "Birds Data Analysis"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(HMSC)
library(vegan)
library(corrplot)
library(cowplot)
select <- dplyr::select
map<- purrr::map
```

# Loading Datasets
```{r message = FALSE}
PA_1970_all <-  read_csv('../BirdData_clean/focalBirds_AllC_1970.csv' )
PA_1970_HQ  <-  read_csv('../BirdData_clean/focalBirds_High_1970.csv' )
PA_2010_all <-  read_csv('../BirdData_clean/focalBirds_AllC_2010.csv' )
PA_2010_HQ  <-  read_csv('../BirdData_clean/focalBirds_High_2010.csv' )

## Object with which species to include
load('../BirdData_clean/finalSP_HQ' )
load('../BirdData_clean/finalSP_all' )
```

# Loading results
```{r}
load('../SavedObjects/BirdVPs/foc_PA_1970_all_both_fracandr2')
load('../SavedObjects/BirdVPs/foc_PA_2010_all_both_fracandr2')
load('../SavedObjects/BirdVPs/foc_PA_1970_HQ_both_fracandr2')
load('../SavedObjects/BirdVPs/foc_PA_2010_HQ_both_fracandr2')
```

# Reading in and filtering results

```{r}

ComputePartitionsFromSegments <- function(x){
  x %>%
    map_dfc(as.data.frame) %>% 
    rownames_to_column(var = 'species') %>%
    set_names(c( "species" , "c", "b", "a", "e", "f", "d", "g")) %>% 
    transmute(species=species,
              env = a + f + 0.5 * d + 0.5 * g,
              env = ifelse(env < 0, 0, env),
              spa = b + e + 0.5 * d + 0.5 * g,
              spa = ifelse(spa < 0, 0, spa),
              codist = c,
              codist = ifelse(codist < 0, 0, codist),
              r2 = env + spa + codist) %>%
    return()
}

```

```{r}
PA_1970_all_both_fracandr2$res %>% 
  ComputePartitionsFromSegments-> PA_1970_all_vpDF

PA_2010_all_both_fracandr2$res %>% 
 ComputePartitionsFromSegments-> PA_2010_all_vpDF

PA_1970_HQ_both_fracandr2$res %>% 
  ComputePartitionsFromSegments-> PA_1970_HQ_vpDF

PA_2010_HQ_both_fracandr2$res %>% 
  ComputePartitionsFromSegments-> PA_2010_HQ_vpDF
```

## Overall Results

```{r}
colMeans(PA_1970_HQ_vpDF[,-1])
colMeans(PA_2010_HQ_vpDF[,-1])

colMeans(PA_1970_all_vpDF[,-1])
colMeans(PA_2010_all_vpDF[,-1])

```

# Occupancy Statistics

## All

```{r}
Occu_1970all <-filter(PA_1970_all, english_name %in% finalSP_all)[,-1]
Occu_2010all <-filter(PA_2010_all, english_name %in% finalSP_all)[,-1]

((colSums(Occu_2010all) - colSums(Occu_1970all))/80*100) %>%
  hist(main = 'distribution of abs percentage change\nin site richness')

((rowSums(Occu_2010all) - rowSums(Occu_1970all))/348*100) %>%
  hist(main = 'distribution of abs percentage change\nin species occupancy')


mean(Occu_2010all !=Occu_1970all)

## very similar total occupancy rates
sum(Occu_1970all)
sum(Occu_2010all)
mean(as.matrix(Occu_1970all))
mean(as.matrix(Occu_2010all))

```

## High Quality

```{r}
Occu_1970_HQ <-filter(PA_1970_HQ, english_name %in% finalSP_HQ)[,-1]
Occu_2010_HQ <-filter(PA_2010_HQ, english_name %in% finalSP_HQ)[,-1]

((colSums(Occu_2010_HQ) - colSums(Occu_1970_HQ))/80*100) %>%
  hist(main = 'distribution of abs percentage change\nin site richness')

((rowSums(Occu_2010_HQ) - rowSums(Occu_1970_HQ))/348*100) %>%
  hist(main = 'distribution of abs percentage change\nin species occupancy')


mean(Occu_2010_HQ !=Occu_1970_HQ)

## small decline in total occupancy rates
sum(Occu_1970_HQ)
sum(Occu_2010_HQ)

mean(as.matrix(Occu_1970_HQ))
mean(as.matrix(Occu_2010_HQ))

(sum(Occu_2010_HQ)-sum(Occu_1970_HQ) )/sum(Occu_1970_HQ) 


```


## Occupancy dissimiliarty

```{r}
OccuTurnover <- data.frame(species = PA_1970_all_vpDF$species,
                           Occupancy1970 = rowSums(Occu_1970all),
                           Occupancy2010 = rowSums( Occu_2010all),
                           Disimilarity = NA)


for( sp in 1:nrow(OccuTurnover)){
  OccuTurnover$Disimilarity[sp] <-   vegdist( bind_rows(Occu_1970all[sp,], Occu_2010all[sp,]), method = 'jaccard')
}

OccuTurnover$Disimilarity %>% hist(main = 'Distribution of Jaccard dissimialrity for each species (all data)')
```


## Which species are driving change?

```{r}
data.frame(species = PA_1970_all_vpDF$species,
           CoDistChange = PA_2010_all_vpDF$codist- PA_1970_all_vpDF$codist  ) %>%
  left_join(OccuTurnover) %>%
  arrange( desc(CoDistChange^2)) -> SpeciesCodistChange

SpeciesCodistChange %>% as_tibble

```

# Looking into fit at species level

Using 1970 HQ as an example

```{r fig.height=12}
load(file = '../SavedObjects/BirdModels/PA_1970_HQ_hmscmodel1')

fittedParams_70HQ <-apply( model$results$estimation$paramX,MARGIN =c(1,2), FUN = mean)

EnvFit<-fittedParams_70HQ[, c( 1:12)]

EnvFit %>%
  as.data.frame() %>%
  rownames_to_column('species') %>%
  select(-Intercept) %>%
  gather('predictor', 'coeff',  - species   ) -> SpeciesEnvAssoc

## These all seem to make a good deal of sense
SpeciesEnvAssoc %>%
  arrange( desc(coeff^2)) %>%
  head(20)

SpeciesEnvAssoc%>%
  ggplot(aes( x = predictor, y = species, fill = coeff ))+
  geom_tile(alpha = 1)+
  scale_fill_gradient2(mid = 0, name = 'Coefficient')+
  scale_y_discrete(limits=rev)+
  theme_minimal()+
  ylab('Species')+
  xlab('Habitat')+
  theme( axis.text.x = element_text(angle = -90, hjust = 0),
         legend.position = 'top', legend.key.width = unit(1, 'cm')) ->Habitat1970

load(file = '../SavedObjects/BirdModels/PA_2010_HQ_hmscmodel1')

fittedParams_10HQ <-apply( model$results$estimation$paramX,MARGIN =c(1,2), FUN = mean)

EnvFit<-fittedParams_10HQ[, c( 1:12)]

EnvFit %>%
  as.data.frame() %>%
  rownames_to_column('species') %>%
  select(-Intercept) %>%
  gather('predictor', 'coeff',  - species   ) -> SpeciesEnvAssoc

## These all seem to make a good deal of sense
SpeciesEnvAssoc %>%
  arrange( desc(coeff^2)) %>%
  head(20)

SpeciesEnvAssoc%>%
  ggplot(aes( x = predictor, y = species, fill = coeff ))+
  geom_tile(alpha = 1)+
  scale_fill_gradient2(mid = 0, name = 'Coefficient')+
  scale_y_discrete(limits=rev)+
  theme_minimal()+
  ylab('Species')+
  xlab('Habitat')+
  theme( axis.text.x = element_text(angle = -90, hjust = 0),
         legend.position = 'top', legend.key.width = unit(1, 'cm')) -> Habitat2010


plot_grid(Habitat1970,Habitat2010, labels = c('a) 1970' , 'b) 2010'))



ggsave( 'BirdImages/BirdHabitatbothHQ.png', height = 14, width = 12, dpi = 500)

```

# Species Associations

## 1970 HQ
```{r fig.height = 12, fig.width=12}
Array_Omega <-corRandomEff(model, cor = TRUE, burning = FALSE)[,,,1] ##  [sp1, sp2, mcmc slice, latentname]
Mean_Omega <- apply(Array_Omega, MARGIN = c(1,2), mean)

corrplot(corr =Mean_Omega,order = 'hclust', tl.col = 1)


png( 'BirdImages/SpecAssoc_1970_HQ.png', width =15,
     height =15, units = 'in', res = 500 )

corrplot(corr =Mean_Omega,order = 'hclust', tl.col = 1)
dev.off()

```


## 2010 HQ 


```{r fig.height = 12, fig.width=12}

load(file = '../SavedObjects/BirdModels/PA_2010_HQ_hmscmodel1')

Array_Omega <-corRandomEff(model, cor = TRUE, burning = FALSE)[,,,1] ##  [sp1, sp2, mcmc slice, latentname]
Mean_Omega <- apply(Array_Omega, MARGIN = c(1,2), mean)

corrplot(corr =Mean_Omega,order = 'hclust', tl.col = 1)


png( 'BirdImages/SpecAssoc_2010_HQ.png', width =15,
     height =15, units = 'in', res = 500 )

corrplot(corr =Mean_Omega,order = 'hclust', tl.col = 1)
dev.off()

```

# Convergence checking

Just including the larger models here. 

```{r}
library(coda)

SpreadArray <- function( i, Array_Omega){
  XX<- Array_Omega[,,i,1]
  return(as.data.frame(t(as.matrix(XX[lower.tri(XX)], nrow =1))))
}
```

## Function to extract and calcualte diagnositics. 
```{r}
CalcGelman <- function(dataset){
  load(paste0('../SavedObjects/BirdModels/PA_',dataset,'_hmscmodel1'))
  load(paste0('../SavedObjects/BirdModels/PA_',dataset,'_hmscmodel2'))
  load(paste0('../SavedObjects/BirdModels/PA_',dataset,'_hmscmodel3'))
  load(paste0('../SavedObjects/BirdModels/PA_',dataset,'_hmscmodel4'))
  
  ### Main predictors
  coda1_param <- as.mcmc(model, parameters = "paramX")
  coda2_param <- as.mcmc(model2, parameters = "paramX")
  coda3_param <- as.mcmc(model3, parameters = "paramX")
  coda4_param <- as.mcmc(model4, parameters = "paramX")
  
  coda_list = mcmc.list(coda1_param,coda2_param , coda3_param, coda4_param)
  Gelmans<- gelman.diag(coda_list, multivariate = FALSE)
  
  ## Convergence of latents
  Array_Omega1 <-corRandomEff(model, cor = TRUE, burning = FALSE) ##  [sp1, sp2, mcmc slice, latentname]
  Array_Omega2 <-corRandomEff(model2, cor = TRUE, burning = FALSE) ##  [sp1, sp2, mcmc slice, latentname]
  Array_Omega3 <-corRandomEff(model3, cor = TRUE, burning = FALSE) ##  [sp1, sp2, mcmc slice, latentname]
  Array_Omega4 <-corRandomEff(model4, cor = TRUE, burning = FALSE) ##  [sp1, sp2, mcmc slice, latentname]
  
  MCMC_length = dim(Array_Omega1)[3]
  
  Omega_MCMC1<- map_dfr(1:MCMC_length, SpreadArray, Array_Omega1)
  Omega_MCMC2<- map_dfr(1:MCMC_length, SpreadArray, Array_Omega2)
  Omega_MCMC3<- map_dfr(1:MCMC_length, SpreadArray, Array_Omega3)
  Omega_MCMC4<- map_dfr(1:MCMC_length, SpreadArray, Array_Omega4)
  
  Omega_MCMC_list <- mcmc.list(as.mcmc(Omega_MCMC1),as.mcmc(Omega_MCMC2),
                               as.mcmc(Omega_MCMC3), as.mcmc(Omega_MCMC4))
  
  Gelmans_omega<- gelman.diag(Omega_MCMC_list, multivariate = FALSE, transform = TRUE)
  
  Omega_Gelmans<- as.data.frame(Gelmans_omega$psrf)
  Omega_Gelmans$dataset = dataset
  
  Main_gelmans <- as.data.frame(Gelmans$psrf) 
  Main_gelmans$dataset = dataset
  
  return( list( Main_gelmans, Omega_Gelmans))
}

```


```{r eval = FALSE}
CG_2010_HQ  <- CalcGelman('2010_HQ')
CG_2010_all <- CalcGelman('2010_all')
CG_1970_HQ  <- CalcGelman('1970_HQ')
CG_1970_all <- CalcGelman('1970_all')

Main_Diags<- bind_rows(CG_2010_HQ[[1]],  
                       CG_2010_all[[1]] ,
                       CG_1970_HQ[[1]] , 
                       CG_1970_all[[1]] )


Omega_Diags<- bind_rows(CG_2010_HQ[[2]],  
                        CG_2010_all[[2]] ,
                        CG_1970_HQ[[2]] , 
                        CG_1970_all[[2]] )

## takes a few minutes so pre-saving

write_csv(Main_Diags, '../SavedObjects/BirdModels/GRDiags_main.csv' )
write_csv(Omega_Diags, '../SavedObjects/BirdModels/GRDiags_Omega.csv' )
```


```{r}
Main_Diags<-read_csv( '../SavedObjects/BirdModels/GRDiags_main.csv' )
Omega_Diags<-read_csv( '../SavedObjects/BirdModels/GRDiags_Omega.csv' )


Main_Diags %>%
  arrange(desc(`Point est.`)) %>%
  head()

Omega_Diags %>%
  arrange(desc(`Point est.`)) %>%
  head()

Main_Diags %>%
  ggplot( aes( x= `Point est.`))+
  geom_histogram(bins = 50)+
  theme_minimal()+
  facet_grid(.~dataset) +
  geom_vline( xintercept = 1.1, linetype= 3)+
    geom_vline( xintercept =1.2, linetype= 2)+
  theme_minimal()+
  xlab('Point Estimate of Gelman-Rubin Diagnostic') -> MainEffects_plot


Omega_Diags %>%
  ggplot( aes( x= `Point est.`))+
  geom_histogram(bins = 50)+
  theme_minimal()+
  facet_grid(.~dataset) +
  geom_vline( xintercept = 1.1, linetype= 3)+
    geom_vline( xintercept =1.2, linetype= 2)+
  theme_minimal()+
  xlab('Point Estimate of Gelman-Rubin Diagnostic') -> Omega_plot




plot_grid(MainEffects_plot, Omega_plot, nrow = 2,
          labels = c('a) Main Effects',
                     'b) Omega Entries' ), scale = 0.9)

ggsave('BirdImages/ModelCovergence.png', 
       width = 10, height = 5, dpi = 500)

```



