---
title: "Simulation Tests"
author: "Chris Terry"
output: 
  html_document: 
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r warning=FALSE, message = FALSE}
library(tidyverse)
select<-dplyr::select
```

# Introduction

This document defines the parameters used in the main simulations, and conducts the post-simulation analysis. 

The simulation functions are run by `BashScripts/TrialLatin.R`. This was run on the QMUL HPC so includes some global variables from that to control the RunID. 

# Model parameters

```{r}
TrialName="TrialLatin"

set.seed(1)

## 6 levels of each of 4 variables, 1 repeat of each. (2592 replicates)

InputParameters <- expand.grid(Disp= 10^seq(-5,  -2, length = 6),  
                          comp_str= seq(0.2,  1.1, length = 6),
                          R_max = seq(5, 50, length = 6), 
                          EnvNoiseSigma = 10^seq(-3, -1 , length = 6),
                          Repeat=c('A', 'B'))
NumRuns = nrow(InputParameters)

InputParameters$RunID = 1:NumRuns
InputParameters$RunName = TrialName
InputParameters$S_global = 50
InputParameters$Occ_thresh = 0.1
InputParameters$comp_conn= 0.01

write_csv(InputParameters, paste0('../Parameters/' ,TrialName, '.csv' ))
```

# Reading in Data

```{r eval = FALSE}
Outs<-list.files(path = paste0('../SavedObjects/', TrialName, '_VP'), full.names = TRUE)
AllRuns<-map_df( Outs,read_csv, show_col_types = FALSE)
write_csv(AllRuns, '../SavedObjects/LatinSimulation.csv')
```

```{r}
InputParameters <- read_csv(paste0('../Parameters/' ,TrialName, '.csv' ))
AllRuns <- read_csv( '../SavedObjects/LatinSimulation.csv')
AllRuns%>%
    set_names(c( "c", "b", "a", "e", "f", "d", "g", "Sample", "species", "prevalance", 'Occ_thresh', 'RunID')) %>%
    transmute(RunID=RunID, species = species,
              Sample = Sample, prevalance=prevalance, 
              env = a + f + 0.5 * d + 0.5 * g,
              env = ifelse(env < 0, 0, env),
              spa = b + e + 0.5 * d + 0.5 * g,
              spa = ifelse(spa < 0, 0, spa),
              codist = c,
              codist = ifelse(codist < 0, 0, codist),
              r2 = env + spa + codist) -> VP_ALL

VP_ALL %>%
  group_by(RunID, Sample)%>%
  mutate(Sample = if_else(Sample=='Before_WithTempVar', 'Before', 'During')) %>%
  summarise( Mean_env = mean(env),
             Mean_spa = mean(spa),
             Mean_codist = mean(codist),
             Mean_r2 = mean(r2)) %>%
  pivot_wider(id_cols = RunID ,
              names_from = Sample , 
              values_from =  starts_with('Mean_')) %>%
  mutate( Shift_Env  =Mean_env_During-Mean_env_Before ,
          Shift_Spa =Mean_spa_During-Mean_spa_Before ,
          Shift_CoDist = Mean_codist_During-Mean_codist_Before,
          Shift_r2 = Mean_r2_During - Mean_r2_Before) %>%
  left_join(InputParameters, by = "RunID") -> AllData

```


# Main results

```{r}
# AllData %>%
#   select( RunID, starts_with('Shift_')) %>%
#   gather( 'Aspect', 'Fraction', starts_with('Shift_'))%>%
#   ggplot( aes( y=Fraction, x = Aspect, fill = Aspect ))+
#   geom_boxplot()+
#   ylab(bquote('Shift in Variance Explained'~(ΔR[adjust]^2)))+
#   xlab('Partition')+
#   scale_x_discrete(  limits = c('Shift_CoDist', 'Shift_Env' ,'Shift_Spa' , 'Shift_r2'),
#                     labels = c('Codistribution','Environmental' ,'Spatial', 'Total'))+
#   scale_fill_manual( values = c('lightskyblue3', 'palegreen4','gold2',  'pink'),
#                      limits = c('Shift_CoDist', 'Shift_Env' ,'Shift_Spa' , 'Shift_r2'))+
#    geom_hline(yintercept = 0, linetype = 2, size = 2)+
#   guides( fill = FALSE)+
#   coord_cartesian(ylim = c( -0.15, 0.2) )+
#   theme_minimal()

# ggsave( 'Figures/SummaryShifts.png', height = 5, width = 6, dpi = 500, bg = 'white')

AllData %>%
  select( RunID, starts_with('Shift_')) %>%
  gather( 'Aspect', 'Fraction', starts_with('Shift_')) %>%
  ungroup( ) %>%
  group_by(Aspect) %>%
  summarise( min(Fraction), max (Fraction))
```


# Impact of model parameters on 'before' communities

```{r}
AllData %>%
  mutate(EnvNoiseSigma = log10(EnvNoiseSigma),
         Disp = log10(Disp)) -> AllData_logg


AllData_logg %>%
  select( RunID, ends_with('_Before'), Disp:EnvNoiseSigma) %>%
  rename( Environmental  =  Mean_env_Before,
          Spatial        =  Mean_spa_Before,
          Codistribution =  Mean_codist_Before,
          Total          =  Mean_r2_Before ) %>%
  gather( 'Aspect', 'Fraction',
          Environmental ,Spatial,Codistribution,Total         ) %>%
  ungroup() %>%
  select(-RunID) %>%
  rename(  'Dispersal (log delta)' = Disp , 'Competition (alpha)'  =comp_str ,
         'Growth (gamma)' = R_max,'Variability (log sigma)' = EnvNoiseSigma ) %>%
  gather( 'InputParameter', 'Value',   -c(Aspect,Fraction) ) %>%
  mutate( Value= signif(Value,2)) %>%
  ggplot( aes( y=Fraction,  x=factor(Value) ))+
  #geom_point(aes(col = Aspect), alpha = 0.5, size = 0.6)+
 # geom_smooth( col = 'black', se = FALSE)+
  geom_boxplot(aes(fill = Aspect))+
  facet_grid(Aspect~InputParameter, scales = 'free')+
  #ggtitle( 'Effect of Parameters Rate on Before Structure')+
  xlab('Model Parameter Value\n(NB Dispersal and Variability on log10 scales)')+
  ylab('R2 Partition\nbefore envrionemntal change')+
  theme_minimal()+
  guides( fill = FALSE)+
  scale_fill_manual( values = c('lightskyblue3', 'palegreen4','gold2',  'pink'),
                       limits = c('Codistribution', 'Environmental' ,
                                  'Spatial' , 'Total'))

# ggsave( 'Figures/ParamStart.png', height = 10, width = 10, dpi = 500)

```


# Impact of model parameters on observed shifts

## Statistics

```{r}
### Main effects, no interactions
lm(data =AllData_logg, Shift_CoDist ~scale(Disp)+scale(comp_str)+scale(R_max)+scale(EnvNoiseSigma)) %>% summary
lm(data =AllData_logg, Shift_Spa    ~scale(Disp)+scale(comp_str)+scale(R_max)+scale(EnvNoiseSigma)) %>% summary
lm(data =AllData_logg, Shift_Env    ~scale(Disp)+scale(comp_str)+scale(R_max)+scale(EnvNoiseSigma)) %>% summary
lm(data =AllData_logg, Shift_r2     ~scale(Disp)+scale(comp_str)+scale(R_max)+scale(EnvNoiseSigma)) %>% summary

### Main effects, 1st order interactions

lm(data =AllData_logg, Shift_CoDist ~(scale(Disp)+scale(comp_str)+scale(R_max)+scale(EnvNoiseSigma))^2) %>% summary
```

## Plots

```{r}
AllData_logg%>%
  select( RunID, starts_with('Shift_'), Disp:EnvNoiseSigma) %>%
  rename( Environmental  =  Shift_Env ,
          Spatial        =  Shift_Spa ,
          Codistribution =  Shift_CoDist ,
          Total          =  Shift_r2     ) %>%
  gather( 'Aspect', 'Fraction',
          Environmental ,Spatial,Codistribution,Total         ) %>%
  ungroup() %>%
select(  -RunID) %>%
  rename(  'Dispersal (log delta)' = Disp , 'Competition (alpha)'  =comp_str ,
         'Growth (gamma)' = R_max,'Variability (log sigma)' = EnvNoiseSigma ) %>%
  gather( 'InputParameter', 'Value',   -c(Aspect,Fraction) ) %>%
  ggplot( aes( y=Fraction,  x=Value ))+
  geom_boxplot(aes(fill = Aspect, x=factor(Value) ))+
  #geom_point(aes(col = Aspect), alpha = 0.5, size = 0.6)+
#  geom_smooth( col = 'black', se = FALSE)+
  geom_hline(yintercept = 0, col = 'black', linetype = 2)+
  facet_grid(Aspect~InputParameter, scales = 'free')+
#  ggtitle( 'Effect of Parameters on Shifts ')+
  xlab('Model Parameter Value\n(NB Dispersal and Variability on log10 scales)')+
  ylab('Absolute Shift in R2 Partition\nwith envrionemntal change')+
  theme_minimal()+
  guides( fill = FALSE)+
  coord_cartesian(ylim = c( -0.15, 0.15) )+
  scale_fill_manual( values = c('lightskyblue3', 'palegreen4','gold2',  'pink'),
                       limits = c('Codistribution', 'Environmental' ,
                                  'Spatial' , 'Total'))


# ggsave( 'Figures/ParamShift.png', height = 10, width = 10, dpi = 500, bg = 'white')


```

# Trends in occupancy and turnover 

```{r}
VP_ALL %>%
  group_by(RunID,Sample) %>%
  summarise( `SD Occupancy` = sd(prevalance),
             `Mean Occupancy` = mean(prevalance),
             `Total Species` = n()) %>%
  left_join(InputParameters, by = "RunID")-> Run_species_data


Run_species_data %>%
   mutate(EnvNoiseSigma = log10(EnvNoiseSigma),
         Disp = log10(Disp)) %>%
  filter( Sample       == 'Before_WithTempVar') %>%
    rename(  'Dispersal (log delta)' = Disp , 'Competition (alpha)'  =comp_str ,
         'Growth (gamma)' = R_max,'Variability (log sigma)' = EnvNoiseSigma ) %>%
  gather( 'InputParameter', 'Value',
        `Dispersal (log delta)` , `Competition (alpha)`,
         `Growth (gamma)` ,`Variability (log sigma)` )%>%
   gather( 'Response', 'Value2',
         `SD Occupancy` ,`Mean Occupancy`, `Total Species` ) %>%
  ggplot( aes( x = factor(Value), y = Value2))+
  geom_boxplot()+
  facet_grid(Response~ InputParameter, scales = 'free')+
  xlab('Model Parameter Value\n(NB Dispersal and Variability on log10 scales)')+
  ylab('Value before Envrionmental Change')+
  theme_minimal()

# ggsave( 'Figures/OccuBefore.png', height = 6, width = 8, dpi = 500)

```


```{r eval = FALSE}
### Function to pull out the turnover (measured as a Jaccard dissimilarity) between sample time slices

ExtractMeanOccu <- function( RunID, InputParameters){
  
  load(file = paste0('../SavedObjects/', InputParameters$RunName[1], '/Run_',RunID,'_MCobject_PreProcess'))
  
  Side_length = Trial$Region_Length 
  Mat_to_keep <- matrix(TRUE, nrow = Side_length, ncol =Side_length)
  Mat_to_keep[c(1,2,Side_length-1, Side_length), ]<-FALSE
  Mat_to_keep[,c(1,2,Side_length-1, Side_length) ]<-FALSE
  sites_to_keep <-which(Mat_to_keep )
  
  Before_focalsites <- Trial$Before_Matrix[,sites_to_keep]  # originally rows = species, cols = sites 
  During_focalsites <- Trial$During_Matrix[,sites_to_keep]  # originally rows = species, cols = sites 
  
  ## Binarising
  Occ_thresh = InputParameters$Occ_thresh[RunID]
  Before_focalsites <- Before_focalsites> Occ_thresh
  During_focalsites <- During_focalsites> Occ_thresh
  
  ### Then find species that can be straight dropped if they too rarely occur 
  Occupancy_before <- rowSums(Before_focalsites)
  Occupancy_during <- rowSums(During_focalsites)
  
  KeepSpecies <- Occupancy_before>=20 & Occupancy_during>=20
  ## also do a final transpose so have sites as rows, and species as columns
  Before_focalsites_focalspecies_t <- t(Before_focalsites[KeepSpecies, ])*1
  During_focalsites_focalspecies_t <- t(During_focalsites[KeepSpecies, ])*1
#  cat(RunID)
  
  ## Overall turnover
  rbind(   matrix(Before_focalsites_focalspecies_t, nrow=1), 
               matrix(During_focalsites_focalspecies_t, nrow=1) ) -> YY
  

  return(data.frame( RunID  = RunID,
                     N_Sp = ncol( Before_focalsites_focalspecies_t),
                     OverallJaccDiss = vegan::vegdist(YY, method = 'jaccard')[1],
                     BeforeMeanOccu = mean(Before_focalsites_focalspecies_t),
                     DuringMeanOccu = mean(During_focalsites_focalspecies_t)))
}

OccuChange<-map_df(.x = 1:2000,  .f = ExtractMeanOccu,
                   InputParameters=InputParameters )
write_csv(OccuChange,'../SavedObjects/OccuChange_simul.csv')
```

```{r}
OccuChange<- read_csv('../SavedObjects/OccuChange_simul.csv')

OccuChange %>%
  mutate( PercChange= 100* (DuringMeanOccu- BeforeMeanOccu)/BeforeMeanOccu) ->OccuChange

OccuChange%>%
  ggplot( aes ( x=PercChange))+
  geom_histogram()+
  geom_vline(xintercept = mean(OccuChange$PercChange))

mean(OccuChange$BeforeMeanOccu)
mean(OccuChange$DuringMeanOccu)

sd(OccuChange$BeforeMeanOccu*100)
sd(OccuChange$DuringMeanOccu*100)

hist(OccuChange$OverallJaccDiss)
mean(OccuChange$OverallJaccDiss)
sd(OccuChange$OverallJaccDiss)

```

# Session Info

```{r}
sessionInfo()
```
