---
title: "Example Simulation"
author: "Chris Terry"
output: 
  html_document: 
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 Scanning the movement through the ternary plot of a community before and during climate change
 
 First thing that is needed is the presence / absences through time. 

```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(igraph)
library(ggtern)
library(cowplot)

walk( list.files('../Scripts/', full.names = TRUE), source)

```

# Running an example simulation

## Parameters 
```{r}
NumRuns = 1
TrialName="TrackThroughTime"

## numbers were drawn randomly 
InputParameters <- data.frame(RunID = 1, 
                     RunName = TrialName,
                     S_global = 50,
                     Disp= 0.0027807184476405385,  
                     comp_str= 0.3101610479876399,
                     comp_conn= 0.01,  
                     R_max = 10.4629793507047)

# write_csv(InputParameters, paste0('../Parameters/' ,TrialName, '.csv' ))
# 
# dir.create(paste0('../SavedObjects/',TrialName), showWarnings = FALSE)
# dir.create(paste0('../SavedObjects/', TrialName, '_VP'), showWarnings = FALSE)

RunID = 1

## Hardcoded parameters 
num_blocks = 1000 ## total time
Ext_Thresh = 0.00001 
num_steps_per_block = 10 # number of time intervals to advance model per block
ArrivalPop = 0.0001  ## initial density of newly arrived species (made a lower to account for wider initial appearance. )
ImmFrac = 1/250 ## proportion of global species pool that arrive each step (basically 2 each step)
K = 10 # carrying capacity parameter 
Length_CC = 25
Region_Length = 14

Disp     <- InputParameters$Disp[RunID]
comp_str <- InputParameters$comp_str[RunID]
R_max    <- InputParameters$R_max[RunID]
comp_conn    <- InputParameters$comp_conn[RunID]
S_global <- InputParameters$S_global[RunID]

Before_During_Tester<- InitiateSystem( Pars = list( seed =  RunID ,
                                                    Region_Length = Region_Length, # Size of Square of nodes
                                                    S_global = S_global, # number of species in global pool
                                                    comp_conn = comp_conn,  #  competitive connectance
                                                    mutu_conn = NULL,  ## positive interaction 
                                                    mutu_relstr = 0))  

B_MAT   <- Before_During_Tester$B_MAT 
CompMat <- Before_During_Tester$CompMat
DispMat <- Before_During_Tester$DispMat
E_nodes <- Before_During_Tester$E_nodes
EnvOpt  <- Before_During_Tester$EnvOpt
E_nodes_2 <- Before_During_Tester$E_nodes_2
EnvOpt_2  <- Before_During_Tester$EnvOpt_2

R_Mat_2 <- Before_During_Tester$R_Mat_2
R_Mat_3 <- Before_During_Tester$R_Mat_3


### Derived Parameters
timescale = 1/ num_steps_per_block  ## ham fisted adjustor
N_sp <- nrow( B_MAT)
t_max = num_steps_per_block +1
N_Nodes <-  ncol(B_MAT)

### Determining climate change
E_change_noise =c( rep(0, num_blocks-Length_CC),  ## initial period of no noise to build community in 
                   seq(0, 0.25,length.out = Length_CC)) + 
  c( rep(0, 500), 
     rnorm(num_blocks-500, sd = 0.01))
```

## Building the communities

```{r eval = FALSE}
#### run model
#####
## Output results container   array time / species / nodes 
## save 50 before, then 25 during. 

N_t_to_Save = 75

ArrayContainer <- array(NA, dim = c(nrow(B_MAT), ncol(B_MAT), N_t_to_Save ))
Total_species = rep(NA, num_blocks)

for(t in 1:num_blocks){  ## burn-in loop (no saving)
  
  #########
  R_Mat <- Calc_RMat_3(E_nodes, EnvOpt,  E_change=E_change_noise[t],
                       R_max = R_max, R_Mat_2, R_Mat_3)
  ############
  ## Dispersal
  B_MAT <- B_MAT +  TotalDispersal(Disp, B_MAT, DispMat,ImmFrac, N_sp, N_Nodes , ArrivalPop)
  ############
  ## Run Dynamics
  B_MAT <- Dynamics_Discrete_Sparse(B_MAT , t_max, timescale,R_Mat,comp_str, CompMat, K)
  B_MAT <- as.matrix(B_MAT)
  #######
  ## Remove locally extinct species 
  B_MAT[rowSums(B_MAT>Ext_Thresh)==0, ] <- 0  ## This option removes species below threshold in all sites
  #################
  Total_species[t]<- sum(rowSums(B_MAT>Ext_Thresh)>0)
  if(t%%100==0){print(t)}
  if( t > num_blocks-N_t_to_Save){
    ArrayContainer[,,t-(num_blocks-N_t_to_Save)] <- B_MAT
  }
}

Trial <- list(  Name = paste0('Run',InputParameters$RunName[1],'_',RunID),
                ArrayContainer=ArrayContainer,
                ModelSetUp = Before_During_Tester,
                E_change_noise = E_change_noise,
                Total_species= Total_species,
                Region_Length = Region_Length)

# save(Trial, file = paste0('../SavedObjects/', InputParameters$RunName[1], '/Run_',RunID,'_ArrayOutput'))

```

```{r}
load(file = paste0('../SavedObjects/', InputParameters$RunName[1], '/Run_',RunID,'_ArrayOutput'))
ArrayContainer <- Trial$ArrayContainer
```


## Fitting HMSC

NB this code also tests different the threshold to determine if the species is present or not. This wasn't included in the final paper.

```{r eval = FALSE, include = FALSE, echo = FALSE}

TrialName="TrackThroughTime"

.libPaths("~/LibraryCentos/")

sessionInfo()

Start<- Sys.time()
print('Starting at:')
Start

## Load packages
library(tidyverse)
library(igraph)
library(HMSC)
select <- dplyr::select

cat(getwd())
setwd('~/GitHub/MC_coherence/BashScripts/')

### Load all functions
walk( list.files('../Scripts/', full.names = TRUE), source)

RunID =  1

t_test = as.numeric(Sys.getenv("SGE_TASK_ID"))

InputParameters<- read_csv(paste0('../Parameters/',TrialName, '.csv'),
                           show_col_types = FALSE)

load(file = paste0('../SavedObjects/', InputParameters$RunName[1], '/Run_',RunID,'_ArrayOutput'))

for(Occ_thresh in c(0.01, 0.1, 1)  ){

  #########
  # Processing to get focal species, binareised from focal nodes
    Side_length = Trial$Region_Length 

  Mat_to_keep <- matrix(TRUE, nrow = Side_length, ncol =Side_length)
  
  Mat_to_keep[c(1,2,Side_length-1, Side_length), ]<-FALSE
  Mat_to_keep[,c(1,2,Side_length-1, Side_length) ]<-FALSE
  
  sites_to_keep <-which(Mat_to_keep )
  
  
  slice <- Trial$ArrayContainer[,,t_test]
  slice_focalsites <- slice[,sites_to_keep]  # originally rows = species, cols = sites 

  ## Binarising
  slice_focalsites <- slice_focalsites> Occ_thresh

  ### Then find species that can be straight dropped if they too rarely occur 
  
  focal_sites_allslices <- Trial$ArrayContainer[,sites_to_keep,]
  
  
    ### need to do some more calculations about which species to keep - need to occur at at least 20 sites accross all time samples
  ## will be a bit lower as more times

  Occupancy_allslice <- apply(focal_sites_allslices>Occ_thresh, MARGIN = c(1,3), sum)  # make a species x time slice occupancy 
KeepSpecies<- apply(Occupancy_allslice>=20, MARGIN = 1, all)
  
  ## also do a final transpose so have sites as rows, and species as columns
  slice_focalsites_focalspecies_t <- t(slice_focalsites[KeepSpecies, ])*1

  ##  Adding species names for back tracking if necessary
  speciesnames <- paste0('Sp_', which(KeepSpecies))
  
  colnames(slice_focalsites_focalspecies_t) <- speciesnames

  ## Adding node names to match original set up 
  nodenames <- paste0('Node_', sites_to_keep)
  rownames(slice_focalsites_focalspecies_t ) <- nodenames

  E_toUse= Trial$ModelSetUp$E_nodes_2[sites_to_keep]

  # End Processing
  #########
  
  ## Prepping envrionment and spatial data for fitting:
  
    MEM <- read.csv(paste0('../SavedObjects/MEMs/MEMof',Side_length,'side_focal',Side_length-4,'.csv'))
  X_formatted <- cbind(scale(E_toUse),
                       scale(E_toUse)^2,
                       MEM[,-1])  # description of sites . direct and squared environment + a moran eigenvector term (dropping node id numbers)
  
  X_groups <- c(rep("env",3),   # fitting three direct environmental terms (i.e. including intercept)
                rep("spa",ncol(MEM)-1))  ## and as many spatial terms as the eigen vector map does (NB name column dropped also)
  
  N_Sites<- (Side_length-4)^2

  ### Function to prepare all the data for HMSC function 
  formData_slice <- as.HMSCdata(Y = slice_focalsites_focalspecies_t,                   
                                 X = X_formatted,  Random = as.factor(1:N_Sites),         
                                 scaleX = TRUE,  interceptX = TRUE)
  

  ## fitting models
  model_slice<-  hmsc(formData_slice, family = "probit",
                       niter = 20000, nburn = 5000, thin = 10 )  # takes a few minutes
  
  ## Variance partitioning
  vpSpp_slice <-   variPart(model_slice,groupX = X_groups, type = "III",  R2adjust = TRUE)  # takes 6x as long... 

  ## Data reformatting and partitioning 
  vpSpp_slice %>% 
    map(as_tibble) %>%
    bind_cols() %>% 
    mutate(Sample = t_test,
           species =colnames(slice_focalsites_focalspecies_t) ) %>% 
    left_join(data.frame(prevalence = colSums(slice_focalsites_focalspecies_t), 
                         species  = colnames(slice_focalsites_focalspecies_t)   ) ,
              by = "species")  %>%
    set_names(c( "c", "b", "a", "e", "f", "d", "g", "Sample", "species", "prevalance")) %>% 
    transmute(species = species,
              Sample = Sample,
              env = a + f + 0.5 * d + 0.5 * g,
              env = ifelse(env < 0, 0, env),
              spa = b + e + 0.5 * d + 0.5 * g,
              spa = ifelse(spa < 0, 0, spa),
              codist = c,
              codist = ifelse(codist < 0, 0, codist),
              r2 = env + spa + codist,
              iteration = paste0('Run_', RunID)) -> VP_Both
  
  VP_Both$Occ_thresh <- Occ_thresh
  
  write_csv(VP_Both, file = paste0('../SavedObjects/', TrialName, '_VP/TimeSlice_',t_test,'_VP_Occ_', Occ_thresh))
  cat(paste0('\nDone VPing:TimeSlice_',t_test,
             'OccupancyThresh:', Occ_thresh,
             'Time:', Sys.time()))
}
  
print('Finishing at:')
Sys.time()

print('Total time:')
Sys.time() - Start

```

## Loading results from HMSC
```{r}
InputParameters <- read_csv( paste0('../Parameters/' ,TrialName ,'.csv'),
                             show_col_types = FALSE) %>%
  mutate( iteration = paste0('Run_',RunID  ))

Outs<-list.files(path = paste0('../SavedObjects/', TrialName, '_VP'), full.names = TRUE)

AllRuns<-map_df( Outs, read_csv, show_col_types = FALSE)
```

# Ternary Plot figure

```{r}
AllRuns %>%
  group_by( Sample, Occ_thresh) %>%
  summarise(n_sp = n_distinct(species),
            MeanCodist = mean(codist),
            MeanR2 = mean(r2),
            MeanSpa = mean(spa),
            MeanEnv = mean(env))-> Grouped_DF
```

```{r}

Grouped_DF %>% 
  filter( Occ_thresh == 0.1) -> XXXX

  XXXX$BaseT = c( rep(NA,50) ,seq(0, 0.25,length.out = 25)) 

DataForLinePlot<- gather(XXXX,'Partition', 'Value',
                           MeanCodist, MeanR2, MeanSpa, MeanEnv  ) %>%
  mutate(Time = Sample+925)
  
DataForLinePlot %>%
    ggplot( )+
      geom_point(aes( x = Time, y = Value,col = Partition))+
   # geom_smooth( se = FALSE)+
    geom_vline( xintercept = 50+925, linetype='dashed')+
    geom_vline( xintercept = 1000, linetype='dashed')+
     geom_point(aes( x = Time, y = Value,  col = Partition),
                data = filter(DataForLinePlot, Time %in% c(975, 1000)),
             size = 1.5 ,shape = 21, col = 1, stroke = 1.5)+
 theme_minimal()+
    ylab('Variance Explained')+
    xlab('Time')+
    scale_colour_manual( values = c('lightskyblue3', 'palegreen4','gold2',  'pink'),
                     labels = c('Codistribution','Environmental' ,'Spatial', 'Total'),
                     limits = c('MeanCodist', 'MeanEnv' ,'MeanSpa' , 'MeanR2') )+
   guides()+
    theme(legend.position = 'bottom')-> LinePlot
  
  LinePlot

  ggsave('Figures/ExampleLinePlot.png',LinePlot,
         width = 5.5, height = 5, dpi = 500 )
  
  XXXX %>%
    arrange( Sample)%>%
    ggtern(aes(x = MeanEnv, z = MeanSpa, y = MeanCodist, size = MeanR2)) +
    geom_path( alpha = 0.8, size = 0.5,aes(color =BaseT  )) +
    geom_point(aes(color =BaseT  ), alpha = 1, size = 3) +
    scale_T_continuous(limits=c(0,1),breaks=seq(0,1,by=0.1),
                       labels=seq(0,1,by=0.1)) +
    scale_L_continuous(limits=c(0.0,1),breaks=seq(0,1,by=0.1),
                       labels=seq(0,1,by=0.1)) +
    scale_R_continuous(limits=c(0,1),breaks=seq(0,1,by=0.1),
                       labels=seq(0,1,by=0.1)) +
    labs(title = '',
         x = "", xarrow = "Environment",
         y = "",yarrow = "Codistribution",
         z = "", zarrow = "Spatial" ) +
    theme_light() +
    #scale_colour_viridis_c()+
    scale_colour_gradient2(low = 'skyblue',mid = 'blue',
                           high = 'red', midpoint = 0.125,
                           name = 'Environmental\nChange\n')+
    theme_showarrows() +
    scale_size_area(limits = c(0,1), breaks = seq(0,1,0.2)) +
    theme(panel.grid = element_line(color = "darkgrey"),
          axis.title = element_text(size = 8),
          legend.position = 'bottom',
          legend.title  = element_text(size = 8),
          legend.text = element_text(size = 8),
          legend.key.width  = unit(1.5, 'cm'))+
    theme_zoom_R(0.4) -> Ternary
  Ternary
ggsave('Figures/TernaryPlot.png',Ternary,
       width = 6, height = 6, dpi = 500 )



```


# Example of system

```{r}

Time1 <- ArrayContainer[,,49]
Time75 <- ArrayContainer[,,75]


ModelSetUp <- Trial$ModelSetUp

SpSel <- 2

SelectedSp<-data.frame( time1Dens = log10(Time1[SpSel, ]),
                        time75Dens = log10(Time75[SpSel, ]),
                        R1 = t(Calc_RMat(E_nodes = ModelSetUp$E_nodes,
                                       EnvOpt = ModelSetUp$EnvOpt[SpSel],
                                       E_change = 0)),
                        R2 =  ModelSetUp$R_Mat_2[SpSel,],
                        E1 =  ModelSetUp$E_nodes,
                        E2 = ModelSetUp$E_nodes_2,
                        E3 = ModelSetUp$R_Mat_3[SpSel,] ,
                        R_Tot = t(Calc_RMat_3(E_nodes = ModelSetUp$E_nodes,
                                       EnvOpt = ModelSetUp$EnvOpt[SpSel],
                                       E_change = 0, R_Mat_2 =ModelSetUp$R_Mat_2[SpSel,] ,
                                       R_Mat_3 =ModelSetUp$R_Mat_3[SpSel,] )),
                        
                        R_Tot_change = t(Calc_RMat_3(E_nodes = ModelSetUp$E_nodes,
                                       EnvOpt = ModelSetUp$EnvOpt[SpSel],
                                       E_change = 0.25, 
                                       R_Mat_2 =ModelSetUp$R_Mat_2[SpSel,] ,
                                       R_Mat_3 =ModelSetUp$R_Mat_3[SpSel,] ))
                        )

 RL <- Trial$Region_Length 
  
  Node_loc<-  data.frame(NodeID = 1:(RL^2),
                         x=rep((1:RL)/RL, each = RL),
                         y=rep((1:RL)/RL, times = RL))
  
 Mat_to_keep <- matrix(TRUE, nrow = RL, ncol =RL)
  Mat_to_keep[c(1,2,RL-1, RL), ]<-FALSE
  Mat_to_keep[,c(1,2,RL-1, RL) ]<-FALSE
  sites_to_keep <-which(Mat_to_keep )
  
xxx<- as_tibble(bind_cols(SelectedSp, Node_loc))%>% 
  mutate( Kept = (NodeID      %in% sites_to_keep)*1 )


```

### Plots

#### Environment
```{r message = FALSE, warning = FALSE}

theme_nothing <- ggmap::theme_nothing

xxx %>% ggplot(aes(x=x,y =y,fill=E1,
             alpha = Kept))+geom_tile()+
  scale_fill_viridis_c(option = 'B', name = '',
                       limits = c(-0.5,1.5))+
  coord_fixed()+
  theme_nothing(legend = TRUE)+
  guides(alpha = FALSE ) -> plot_E1_start


xxx %>% ggplot(aes(x=x,y =y,fill=E1+0.25,
             alpha = Kept))+geom_tile()+
  scale_fill_viridis_c(option = 'B', name = '',
                       limits = c(-0.5,1.5))+
  coord_fixed()+
  theme_nothing(legend = TRUE)+
  guides(alpha = FALSE , fill=FALSE) -> plot_E1_during

xxx %>% ggplot(aes(x=x,y =y,fill=E2,
             alpha = Kept))+geom_tile()+
  scale_alpha(range = c(0.2, 1))+coord_fixed()+ theme_nothing(legend = TRUE)+
  guides(alpha = FALSE , fill=FALSE)+
    scale_fill_viridis_c(option = 'B', name = '',
                       limits = c(-0.5,1.5))-> plot_E2

xxx %>% 
  ggplot(aes(x=x,y =y,fill=E3,
             alpha = Kept))+scale_alpha(range = c(0.2, 1))+
  geom_tile()+ 
  scale_fill_viridis_c(option = 'B', name = '',
                       limits = c(-0.5,1.5))+
    theme_nothing(legend = TRUE)+
  guides(alpha = FALSE , fill=FALSE)+coord_fixed()-> plot_E3


LegendE <- get_legend(plot_E1_start)

```

#### Growth Rates
```{r}


xxx %>% ggplot(aes(x=x,y =y,fill=R_Tot,
             alpha = Kept))+
  geom_tile()+coord_fixed()+
  scale_alpha(range = c(0.2, 1))+ 
  theme_nothing(legend = TRUE)+guides(alpha ='none')+
  scale_fill_viridis_c(option = 'A', limits = c(0,1), name = 'R') -> plot_Rpre

xxx %>% ggplot(aes(x=x,y =y,fill=R_Tot_change,
                   alpha = Kept))+
  geom_tile()+coord_fixed()+
  scale_alpha(range = c(0.2, 1))+ theme_nothing(legend = TRUE)+
  guides( fill="none", alpha = 'none')+
  scale_fill_viridis_c(option = 'A', limits = c(0,1))-> plot_r_during

LegendR <- get_legend(plot_Rpre)

```

#### Density 


```{r}
xxx %>% 
  ggplot(aes(x=x,y =y,fill=time1Dens,
             alpha = Kept))+
  geom_tile()+scale_fill_viridis_c(limits = c(-10,2),
                                   name = 'Log10\nDensity')+
    theme_nothing(legend = TRUE)+coord_fixed()+
    guides( alpha = 'none')+
  scale_alpha(range = c(0.2, 1))-> plot_dens1


xxx %>% 
  ggplot(aes(x=x,y =y,fill=time75Dens,
             alpha = Kept))+
  scale_alpha(range = c(0.2, 1))+
  geom_tile()+scale_fill_viridis_c()+
  guides( fill="none", alpha = 'none')+
    theme_nothing(legend = TRUE)+coord_fixed() -> plot_dens2


LegendD <- get_legend(plot_dens1)

```

#### Combining

```{r}
plot_grid( plot_E1_start+guides(fill=FALSE),
           plot_E1_during,
           LegendE,
           plot_E2,
           plot_E3,
           LegendE,
           plot_Rpre+guides(fill="none"),
           plot_r_during,
           LegendR,
           plot_dens1+guides(fill="none"),
           plot_dens2,
           LegendD,
           ncol = 3,
           labels = c('a) E1, t=975',
                      'b) E1, t =1000',
                      '',
                      'c) E2', 
                      'd) R* (Sp.X)',
                      '',
                      'e) Sp.X R, t=975',
                      'f) Sp.X R, t =1000',
                      '',
                      'g) Sp.X Density, t=975',
                      'h) Sp.X Density, t =1000',
                      ''),
           hjust = -0.1,
           rel_widths = c(1,1,0.5,1,1,0.5,
                          1,1,0.5,1,1,0.5),
           label_fontface = 'plain',
           scale = 0.9)
                
ggsave( 'Figures/SimulationExample_grid.png', height = 10, width = 5, dpi = 500)

             


```

```{r fig.height=3, fig.width=3}
## Growth Rate components
xxx %>% ggplot(aes(x=x,y =y,fill=R1))+geom_tile()+scale_fill_viridis_c()
xxx %>% ggplot(aes(x=x,y =y,fill=R2))+geom_tile()+scale_fill_viridis_c()
xxx %>% ggplot(aes(x=x,y =y,fill=E3))+geom_tile()+scale_fill_viridis_c()

```


# Example for Main text box


```{r}

plot_grid( plot_E1_start+guides(fill=FALSE),
           plot_E2,
           plot_E3,
           LegendE,
           plot_E1_during,
           plot_dens1+guides(fill="none"),
           plot_dens2,
           LegendD,
           nrow = 2,
           labels = c('a) E1, t=975',
                      'b) E2', 
                      'c) R* (Sp.X)',
                      '',
                      'd) E1, t=1000',
                      'e) Sp.X, t=975',
                      'f) Sp.X, t =1000',
                      ''),
           hjust = -0.1,
           rel_widths = c(1,1,1,0.5, 1,1,0.5),
           label_fontface = 'plain',
           scale = 0.9)
                
ggsave( 'Figures/SimulationExample_MTbox.png', height = 4, width = 6, dpi = 500)

        



```



















 
 
 